Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-08 sections/03-trending.tex
Update on Overleaf.

\begin{table}
    \centering
    \begin{tabularx}{0.75\textwidth}{l X}
        \toprule
        Application & Predicted change computation \\
        \midrule
        Measurement & $(x_{t + l} - x_t)_{t=1}^T \label{eq:diffxl_measure} $\\
        Nowcasting & $
\diffxl =
\begin{cases}
(x_{t+l|t+l} -x_{t|t+l})^T_{t=1} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t+l} -y_{t})^T_{t=1} & \text{else}.
\end{cases} \label{eq:diffxl_nowcasting}
$ \\
        Forecasting & $
\diffxl =
\begin{cases}
(x_{t+l|t} - x_{t|t})^T_{t=1} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t} - y_{t})^T_{t=1}  & \text{else}.
\end{cases} \label{eq:diffxl_forecasting}
$\\
        \bottomrule
    \end{tabularx}
    \caption{Computation of the predicted change in the different applications. Signals $x_{t | \tau}$ refer to values issued at $\tau$ with a target time $t$. Note the different relations between the issue and target time for nowcasts and forecasts. }
    \label{tab:notation}
\end{table}

@@

Analyzing the number of points in the green versus red quadrants is a standard approach in the trending assessment of measurement data \parencite{Critchley2010,Saugel2015}. 
It boils down to assessing the probability of trending $P(\diffxrv \diffyrv > 0)$, where $\diffyrv$ and $\diffxrv$ denote random variables for future incremental changes.
Note that $z_1 z_2 > 0$ if and only if $\sign(z_1) = \sign(z_2)$ ($z_1, z_2 \in \R \setminus \{ 0 \}$).
The standard estimator for $P(\diffxrv \diffyrv > 0)$ is
\begin{equation}
    \acc (\diffx, \diffy) \coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0}}{\card{T}}.\label{eq:acc}
\end{equation}
We refer to this estimator as the trending ratio of the prediction and set $\mathcal{T} = \{1, \dots, T-l\}$.
Visually, the measure computes the fraction of points in the upper right or lower left quadrant.
Such a $2 \times 2$ table of counts is called a contingency table.
It is often used in other scientific areas for evaluation, for example, in dichotomous forecasting or as a confusion matrix in classification analysis~\parencites(see, e.g., the introductions in)()[Ch. 4]{James2021}[Ch. 3]{Jolliffe2012}.
There, a wide range of other methods are usually used to analyze further characteristics of contingency tables.
Two simple measures that focus on specific areas of interest are the positive and negative trending ratios $\accp$ and $\accm$, respectively.
They are defined as
\begin{align}
    \accp (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0}} \label{eq:accp}\\
    \accm (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0}}\label{eq:accm}
\end{align}
In the classification context, these measures are known as positive or negative predictive value and hit rate or detection failure ratio in forecasting.
They give the probability of a correct prediction of the direction of change, given that the signal direction is positive or negative, respectively.
Therefore, they measure the trending ability of the prediction for positive and negative changes separately by cutting the considered data.
In terms of probability, the two measures give estimates of $P(\diffxrv \diffyrv > 0 | \diffxrv > 0)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0)$, respectively.

@@

All measures can also be evaluated as a rolling estimate to detect changes in performance over time.
A rolling estimate is a sequence of estimates, each computed on a length-$w$-window of the data.
For the trending ratio, a rolling estimate with a backward-looking window is given by
\begin{equation*}
    \acc_t (\diffx, \diffy) \coloneqq \frac{\sum_{t^\star = t-w + 1}^{t} \ind{\diffxt[t^\star] \diffyt[t^\star] > 0}}{w}, \quad t = w-1, \dots, T.\label{eq:acc_rolling}
\end{equation*}
Figure~\ref{fig:trending_ratio_time_series} depicts a rolling window estimate of the trending ratio for the simulated data of Figures~\ref{fig:trending_basic_4q_sample} and~\ref{fig:trending_basic_4q_sample_color}.
Thus, the yearly course of the prediction trending ratio can be detected.
The trending ability of the signal has a strong sinus-shaped seasonality with a trending ability peak after a quarter of a year and a low point after three quarters.
This seasonal behavior cannot be distinguished in the two four-quadrant plots or the trending ratio.

@@

If the true value is given as a distribution because it is still unknown, the probabilities $p_t$ can be computed by integration.
Let for two nowcasts the distributions be given by \acp{pdf} $f_{t+l|t+l}$ and $f_{t|t+l}$ with \acp{cdf}  $F_{t+l|t+l}$ and $F_{t|t+l}$.
Then, the probability of a negative change can be computed by
\begin{align}
    \bar{p}_t
        &= \int_{\begin{subarray}{l}x_1, x_2 \in \R: \\ x_2 < x_1\end{subarray}} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: (x_1, x_2) \nonumber\\
        &= \int_{x_1 \in \R} \int_{-\infty}^{x_1} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \nonumber\\
        &= \int_{x_1 \in \R} f_{t|t+l} (x_1) F_{t+l|t+l} (x_1)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \label{eq:trending-probabilistic-pdf}.
\end{align}
\unsure{Die Unabhängigkeitsannahme ist nicht zu streng? weil es ist ja eigentlich die Aufgabe von einenm guten Nowcaster diese Abhängigkeit zu modellieren oder?}
Thereby, the distributions are assumed to be independent.
If the nowcasts have the form of a multivariate distribution, including the dependence of the two \acp{pdf}, $f_{t+l|t+l} (x_2)$ has to be replaced by the \ac{pdf} conditional on $x_1$.
As a Monte Carlo approximation of Equation~\eqref{eq:trending-probabilistic-pdf}, the probability can also be calculated by sampling from $f_{t+l|t+l}$ and $f_{t|t+l}$ and calculating the fraction of negative changes.
For forecasts, the indexes have to be shifted.
If no \acp{pdf} are available, they can be estimated from the \ac{cdf} or quantiles, or the \ac{cdf} or quantiles can be used to generate samples for the Monte Carlo approximation.

>>>>

\begin{table}
    \centering
    \begin{tabularx}{0.75\textwidth}{l X}
        \toprule
        Application & Predicted change computation \\
        \midrule
        Measurement & $(x_{t + l} - x_t)_{t=1}^T \label{eq:diffxl_measure} $\\
        Nowcasting & $
\diffxl =
\begin{cases}
(x_{t+l|t+l} -x_{t|t+l})^T_{t=1} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t+l} -y_{t})^T_{t=1} & \text{else}.
\end{cases} \label{eq:diffxl_nowcasting}
$ \\
        Forecasting & $
\diffxl =
\begin{cases}
(x_{t+l|t} - x_{t|t})^T_{t=1} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t} - y_{t})^T_{t=1}  & \text{else}.
\end{cases} \label{eq:diffxl_forecasting}
$\\
        \bottomrule
    \end{tabularx}
    \caption{Computation of the predicted change in the different applications. Signals $x_{t | \tau}$ refer to values issued at $\tau$ with a target time $t$. Note the different relations between the issue and target time for nowcasts and forecasts. \todo{$T-l$}}
    \label{tab:notation}
\end{table}

@@

Analyzing the number of points in the green versus red quadrants is a standard approach in the trending assessment of measurement data \parencite{Critchley2010,Saugel2015}. 
It boils down to assessing the probability of trending $P(\diffxrv \diffyrv > 0)$, where $\diffyrv$ and $\diffxrv$ denote random variables for future incremental changes.
Note that $z_1 z_2 > 0$ if and only if $\sign(z_1) = \sign(z_2)$ ($z_1, z_2 \in \R \setminus \{ 0 \}$).
The standard estimator for $P(\diffxrv \diffyrv > 0)$ is
\begin{equation}
    \acc (\diffx, \diffy) \coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0}}{\card{T}}.\label{eq:acc}
\end{equation}
We refer to this estimator as the trending ratio of the prediction and set $\mathcal{T} = \{1, \dots, T-l\}$. \todo{Durchwalken; nennen, warum ratio; Anteil concordant points an allen, deswegen nennen wir es ratio}
Visually, the measure computes the fraction of points in the upper right or lower left quadrant.
Such a $2 \times 2$ table of counts is called a contingency table.
It is often used in other scientific areas for evaluation, for example, in dichotomous forecasting or as a confusion matrix in classification analysis~\parencites(see, e.g., the introductions in)()[Ch. 4]{James2021}[Ch. 3]{Jolliffe2012}.
There, a wide range of other methods are usually used to analyze further characteristics of contingency tables.
Two simple measures that focus on specific areas of interest \todo{welche areas} are the positive and negative trending ratios $\accp$ and $\accm$, respectively.
They are defined as
\begin{align}
    \accp (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0}} \label{eq:accp}\\
    \accm (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0}}\label{eq:accm}
\end{align}
In the classification context, these measures are known as positive or negative predictive value and hit rate or detection failure ratio in forecasting.
They give the probability of a correct prediction of the direction of change, given that the signal direction is positive or negative, respectively.
Therefore, they measure the trending ability of the prediction for positive and negative changes separately by cutting the considered data.
In terms of probability, the two measures give estimates of $P(\diffxrv \diffyrv > 0 | \diffxrv > 0)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0)$, respectively.

@@

All measures can also be evaluated as a rolling estimate to detect changes in performance over time. \todo{Klarer, warum man das soll. Wir wollen eigentlich trending ratio jetzt wissen. }
A rolling estimate is a sequence of estimates, each computed on a length-$w$-window of the data.
For the trending ratio, a rolling estimate with a backward-looking \todo{Klarer, was heißt backward-looking} window is given by
\begin{equation*}
    \acc_t (\diffx, \diffy) \coloneqq \frac{\sum_{t^\star = t-w + 1}^{t} \ind{\diffxt[t^\star] \diffyt[t^\star] > 0}}{w}, \quad t = w-1, \dots, T.\label{eq:acc_rolling}
\end{equation*}
\todo{Was heißt w? Großes w weniger fokussiert auf current situation; kleines w lokaler}
Figure~\ref{fig:trending_ratio_time_series} depicts a rolling window estimate of the trending ratio for the simulated data of Figures~\ref{fig:trending_basic_4q_sample} and~\ref{fig:trending_basic_4q_sample_color}.
Thus, the yearly course of the prediction trending ratio can be detected.
The trending ability of the signal has a strong sinus-shaped seasonality with a trending ability peak after a quarter of a year and a low point after three quarters.
This seasonal behavior cannot be distinguished in the two four-quadrant plots or the trending ratio.

@@

If the true value is given as a distribution because it is still unknown, the probabilities $p_t$ can be computed by integration.
Let for two nowcasts the distributions be given by \acp{pdf} $f_{t+l|t+l}$ and $f_{t|t+l}$ with \acp{cdf}  $F_{t+l|t+l}$ and $F_{t|t+l}$.
Then, the probability of a negative change can be computed by
\begin{align}
    \bar{p}_t
        &= \int_{\begin{subarray}{l}x_1, x_2 \in \R: \\ x_2 < x_1\end{subarray}} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: (x_1, x_2) \nonumber\\
        &= \int_{x_1 \in \R} \int_{-\infty}^{x_1} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \nonumber\\
        &= \int_{x_1 \in \R} f_{t|t+l} (x_1) F_{t+l|t+l} (x_1)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \label{eq:trending-probabilistic-pdf}.
\end{align}
Thereby, the distributions are assumed to be independent.
If the nowcasts have the form of a multivariate distribution, including the dependence of the two \acp{pdf}, $f_{t+l|t+l} (x_2)$ has to be replaced by the \ac{pdf} conditional on $x_1$.
As a Monte Carlo approximation of Equation~\eqref{eq:trending-probabilistic-pdf}, the probability can also be calculated by sampling from $f_{t+l|t+l}$ and $f_{t|t+l}$ and calculating the fraction of negative changes.
For forecasts, the indexes have to be shifted.
If no \acp{pdf} are available, they can be estimated from the \ac{cdf} or quantiles, or the \ac{cdf} or quantiles can be used to generate samples for the Monte Carlo approximation.
